---
title: "Topic_2_adverse_outcomes_severity_score"
author: "Uvarova Victoria"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE)

library(tidyverse)
library(broom)
library(performance)
library(ggResidpanel)
library(forestplot)
```

## Loading the Dataset for Analysis

```{r}
sepsis_HA_df_analyze <- readRDS("C:/Users/krine/Downloads/Telegram Desktop/joint_sepsis.rds")
```

## Data Preprocessing

### 1. Categorization of Variables

NLR (Neutrophil-to-Lymphocyte Ratio) [10.1016/j.jtauto.2022.100159]. The categories "Neutropenia" (0.1-0.7) and "Normal range" (0.7-2) were combined into a single category "Normal range" because the "Neutropenia" category includes only 6 patients, which is insufficient for a regression model.

Age (18--44, 45--59, 60--74, 75+)

```{r}
sepsis_HA_df_analyze <- sepsis_HA_df_analyze %>%
  mutate(
    NLR_category = case_when(
      is.na(NLR) ~ NA_character_,
      # NLR >= 0.1 & NLR <= 0.7 ~ "Neutropenia",
      NLR >= 0.1 & NLR <= 2 ~ "Normal range",
      NLR > 2 & NLR <= 3 ~ "Grey zone",
      NLR > 3 & NLR <= 7 ~ "Mild to moderate inflammation",
      NLR > 7 & NLR <= 11 ~ "Moderate to severe inflammation",
      NLR > 11 & NLR <= 17 ~ "Severe inflammation",
      NLR > 17 & NLR <= 23 ~ "Critical inflammation: septic shock",
      NLR > 23 ~ "Critical inflammation: politrauma, cancer, surgery"
    )
  ) %>%
  
  mutate(NLR_category = factor(
    NLR_category,
    levels = c(
      "Neutropenia",
      "Normal range",
      "Grey zone",
      "Mild to moderate inflammation",
      "Moderate to severe inflammation",
      "Severe inflammation",
      "Critical inflammation: septic shock",
      "Critical inflammation: politrauma, cancer, surgery"
    )
  )) %>%
  
  mutate(
    age_category = case_when(
      age >= 18 & age <= 44 ~ "18–44",
      age >= 45 & age <= 59 ~ "45–59",
      age >= 60 & age <= 74 ~ "60–74",
      age >= 75 ~ "75+"
    )
  ) %>%
  
  mutate(age_category = factor(age_category, levels = c("18–44", "45–59", "60–74", "75+")))
```

### 2. Time-to-Death Stratification

Variables: ICU_dur_T0_first_end (duration): length of ICU stay. ICU_first_end_status (end status): patient's ICU outcome.

Patients who died within 6 days were classified as "early mortality", those who died after 6 days as "late mortality", and those who recovered as "recovery".

```{r}
sepsis_HA_df_analyze <- sepsis_HA_df_analyze %>%
  mutate(
    death_category = case_when(
      ICU_first_end_status == "death" &
        ICU_dur_T0_first_end >= 0 &
        ICU_dur_T0_first_end <= 6 ~ "early mortality",
      ICU_first_end_status == "death" &
        ICU_dur_T0_first_end > 6 ~ "late mortality",
      ICU_first_end_status == "recovery" ~ "recovery"
    )
  ) %>%
  mutate(death_category = factor(
    death_category,
    levels = c("early mortality", "late mortality", "recovery")
  )) 
```

### 3. Histograms of variable distribution

Log transformation is required for: WBC, PLT, creatinine, VIS2020, NEU, LYM, bilirubin_total, NLR, SII, CRP, PCT, shock_dur_T0_first_end, ICU_dur_T0_first_end, MV_dur_T0_first_end, lactate, aPTT; not required for: age, SOFA, PF_ratio, weight, MAP, fibrinogen.

Applying log(Value) without adding a small constant causes errors for zeros and excludes them from the plot, leaving only positive values in the distribution.

```{r, fig.width=9, fig.height=9}
selected_vars <- sepsis_HA_df_analyze %>%
  filter(time_point == 0) %>%
  dplyr::select(
    "age",
    "SOFA",
    "MAP",
    "WBC",
    "PLT",
    "creatinine",
    "VIS2020",
    "NEU",
    "LYM",
    "bilirubin_total",
    "NLR",
    "SII",
    "weight",
    "CRP",
    "PF_ratio",
    "PCT",
    "shock_dur_T0_first_end",
    "ICU_dur_T0_first_end",
    "MV_dur_T0_first_end",
    "lactate",
    "fibrinogen",
    "aPTT"
  ) %>% pivot_longer(cols = everything(),
                     names_to = "Variable",
                     values_to = "Value")
ggplot(selected_vars, aes(x = Value)) +
  geom_histogram() +
  facet_wrap( ~ Variable, scales = "free", ncol = 5) +
  theme_bw() + labs(title = "Distribution of Variables")

selected_vars %>%
  mutate(Value = log(Value)) %>%
  ggplot(aes(x = Value)) +
  geom_histogram() +
  facet_wrap( ~ Variable, scales = "free", ncol = 5) +
  theme_bw() + labs(title = "Distribution of Log-transformed Variables")
```

### 4. Log-transformation

Log-transformation of selected variables with an adaptive epsilon based on 10% of the smallest positive value to handle zeros (data sepsis_HA_df_analyze_log).

```{r, fig.width=9, fig.height=9}
add_epsilon <- function(x, factor = 0.1) {
  epsilon <- min(x[x > 0], na.rm = TRUE) * factor
  log(x + epsilon)
}

sepsis_HA_df_analyze_log <- sepsis_HA_df_analyze %>%
  filter(time_point == 0) %>%
  mutate(across(
    c(
      WBC,
      PLT,
      creatinine,
      VIS2020,
      NEU,
      LYM,
      bilirubin_total,
      NLR,
      SII,
      CRP,
      PCT,
      shock_dur_T0_first_end,
      ICU_dur_T0_first_end,
      MV_dur_T0_first_end,
      lactate,
      aPTT
    ),
    ~ add_epsilon(.)
  ))

sepsis_HA_df_analyze_log_longer <- sepsis_HA_df_analyze_log %>%
  dplyr::select(
    "age",
    "SOFA",
    "MAP",
    "WBC",
    "PLT",
    "creatinine",
    "VIS2020",
    "NEU",
    "LYM",
    "bilirubin_total",
    "NLR",
    "SII",
    "weight",
    "CRP",
    "PF_ratio",
    "PCT",
    "shock_dur_T0_first_end",
    "ICU_dur_T0_first_end",
    "MV_dur_T0_first_end",
    "lactate",
    "fibrinogen",
    "aPTT"
  ) %>%
  pivot_longer(cols = everything(),
               names_to = "Variable",
               values_to = "Value")


ggplot(sepsis_HA_df_analyze_log_longer, aes(x = Value)) +
  geom_histogram() +
  facet_wrap( ~ Variable, scales = "free", ncol = 5) +
  theme_bw() + labs(title = "Distribution of Variables")
```

## Mortality rate

256 patients were admitted to the ICU, with a sepsis mortality rate of 49.2%.

```{r}
death_count <- sum(
  sepsis_HA_df_analyze$ICU_first_end_status == "death" &
    sepsis_HA_df_analyze$time_point == 0
)
total_count <- sum(sepsis_HA_df_analyze$time_point == 0)
mortality_rate <- (death_count / total_count) * 100

cat(
  "Total patients:",
  total_count,
  "\n",
  "Mortality rate:",
  round(mortality_rate, 1),
  "%\n"
)
```

## Predictors of mortality in patients with sepsis [10.1097/MD.0000000000035335]

Univariate and multivariate logistic regression analysis of risk factors associated with mortality in patients with sepsis.

Statistically significant variables in the baseline comparison (unadjusted univariate logistic regression analysis): age, SOFA, PLT, DDimer, PF_ratio, albumin, creatinine, lactate.

Statistically significant variables in multivariate logistic regression analysis (adjusted): age, lactate, PLT (platelet count).

## Univariate analysis

The logistic regression model predicts the probability of the outcome "death" relative to the baseline level "recovery". The forest plot represents the odds ratios (OR) for various predictors of the outcome (ICU_first_end_status) at time point 0, along with their 95% CI. Logistic regression estimates coefficients in terms of log odds. If the coefficient equals 0, then OR = e\^0 = 1 , which corresponds to no effect.

### Variables without log-transformation


```{r, fig.width=8, fig.height=8}
sepsis_HA_df_analyze$ICU_first_end_status <- relevel(sepsis_HA_df_analyze$ICU_first_end_status, ref = "recovery") # baseline category recovery
sepsis_HA_df_analyze$sex <- relevel(sepsis_HA_df_analyze$sex, ref = "female") # baseline category female

# code can be used for any time point
time_points <- unique(sepsis_HA_df_analyze$time_point)

predictors <- c(
  "sex",
  "age",
  "shock_T0",
  "MV_T0",
  "SOFA",
  "MAP",
  "WBC",
  "PLT",
  "creatinine",
  "VIS2020",
  "NEU",
  "LYM",
  "bilirubin_total",
  "NLR",
  "SII",
  "weight",
  "CRP",
  "PF_ratio",
  "PCT",
#  "shock_dur_T0_first_end",
#  "ICU_dur_T0_first_end",
#  "MV_dur_T0_first_end",
  "lactate",
  "fibrinogen",
  "aPTT"
)

all_results <- lapply(time_points, function(tp) {
  data_tp <- sepsis_HA_df_analyze %>% filter(time_point == tp)
  
  results <- lapply(predictors, function(var) {
    formula <- as.formula(paste("ICU_first_end_status ~", var, "+ combined_efferon + study_id"))
    model <- glm(formula, data = data_tp, family = binomial)
    
    summary <- summary(model)$coefficients[2, ]
    conf_int <- confint(model)[2, ]
    
    tibble(
      variable = var,
      estimate = summary[1],
      conf.low = conf_int[1],
      conf.high = conf_int[2],
      p.value = summary[4],
      time_point = tp
    )
  })
  bind_rows(results)
})

all_results_df <- bind_rows(all_results)
all_results_df <- all_results_df %>%
  filter(time_point == 0) %>%
  mutate (p_adj = p.adjust(p.value, method = "BH")) %>%
  mutate(significance = case_when(p_adj < 0.001 ~ "***", p_adj < 0.01 ~ "**", p_adj < 0.05 ~ "*", TRUE ~ ""))

# forest_data_time_point_0
forest_data_time_point_0 <- all_results_df %>%
  mutate(
    odds_ratio = exp(estimate),
    conf.low = exp(conf.low),
    conf.high = exp(conf.high)
  )

label_text <- cbind(
  Variable = forest_data_time_point_0$variable,
  OR = sprintf("%.2f", forest_data_time_point_0$odds_ratio),
  CI = sprintf(
    "[%.2f, %.2f]",
    forest_data_time_point_0$conf.low,
    forest_data_time_point_0$conf.high
  ),
  Significance = forest_data_time_point_0$significance
)

forestplot(
  labeltext = label_text,
  mean = forest_data_time_point_0$odds_ratio,
  lower = forest_data_time_point_0$conf.low,
  upper = forest_data_time_point_0$conf.high,
  is.summary = FALSE,
  xlab = "Odds Ratio (95% CI)",
  zero = 1,
  boxsize = 0.3,
  col = fpColors(
    box = "blue",
    line = "darkblue",
    zero = "red"
  ),
  title = "Forest plot (variables without log-transformation)",
  txt_gp = fpTxtGp(xlab = gpar(fontsize = 20), ticks = gpar(fontsize = 20))
)

```

### Variables with log-transformation

```{r, fig.width=8, fig.height=8}
sepsis_HA_df_analyze_log$ICU_first_end_status <- relevel(sepsis_HA_df_analyze_log$ICU_first_end_status, ref = "recovery") # baseline category recovery
sepsis_HA_df_analyze_log$sex <- relevel(sepsis_HA_df_analyze_log$sex, ref = "female") # baseline category female

predictors <- c(
  "sex",
  "age",
  "shock_T0",
  "MV_T0",
  "SOFA",
  "MAP",
  "WBC",
  "PLT",
  "creatinine",
  "VIS2020",
  "NEU",
  "LYM",
  "bilirubin_total",
  "NLR",
  "SII",
  "weight",
  "CRP",
  "PF_ratio",
  "PCT",
#  "shock_dur_T0_first_end",
#  "ICU_dur_T0_first_end",
#  "MV_dur_T0_first_end",
  "lactate",
  "fibrinogen",
  "aPTT"
)

results_log <- lapply(predictors, function(var) {
  formula <- as.formula(paste("ICU_first_end_status ~", var, "+ combined_efferon + study_id"))
  model <- glm(formula, data = sepsis_HA_df_analyze_log, family = binomial)
  
  summary <- summary(model)$coefficients[2, ]
  conf_int <- confint(model)[2, ]
  
  tibble(
    variable = var,
    estimate = summary[1],
    conf.low = conf_int[1],
    conf.high = conf_int[2],
    p.value = summary[4]
  )
})

results_log_df <- bind_rows(results_log)
results_log_df <- results_log_df %>%
  mutate (p_adj = p.adjust(p.value, method = "BH")) %>%
  mutate(significance = case_when(p_adj < 0.001 ~ "***", p_adj < 0.01 ~ "**", p_adj < 0.05 ~ "*", TRUE ~ "")) %>%
  mutate(
    odds_ratio = exp(estimate),
    conf.low = exp(conf.low),
    conf.high = exp(conf.high)
  )

label_text_log <- cbind(
  Variable = results_log_df$variable,
  OR = sprintf("%.2f", results_log_df$odds_ratio),
  CI = sprintf(
    "[%.2f, %.2f]",
    results_log_df$conf.low,
    results_log_df$conf.high
  ),
  Significance = results_log_df$significance
)

forestplot(
  labeltext = label_text_log,
  mean = results_log_df$odds_ratio,
  lower = results_log_df$conf.low,
  upper = results_log_df$conf.high,
  is.summary = FALSE,
  xlab = "Odds Ratio (95% CI)",
  zero = 1,
  boxsize = 0.3,
  col = fpColors(
    box = "blue",
    line = "darkblue",
    zero = "red"
  ),
  title = "Forest plot (variables with log-transformation)",
  txt_gp = fpTxtGp(xlab = gpar(fontsize = 20), ticks = gpar(fontsize = 20))
)
```

## Multivariate Analysis

### Variables without log-transformation

```{r, fig.width=8, fig.height=8}
sepsis_HA_df_analyze_time_point_0 <- sepsis_HA_df_analyze %>%
  filter(time_point == 0)

multivariate_model <- glm(
  ICU_first_end_status ~ combined_efferon + sex + age + MV_T0 + SOFA + VIS2020 + lactate,
  data = sepsis_HA_df_analyze_time_point_0,
  family = binomial
)

check_model(multivariate_model)
```

