---
title: "Treatment impact on parameters"
author: "Anastasiia Potamoshneva"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(broom)
library(ggfortify)
library(ggResidpanel)
library(performance)
library(lmtest)
library(sandwich)
library(car)
library(DT)
library(forestplot)
```

## Data loading

```{r}
sepsis <- readRDS("C:\\Users\\user\\RStudio_projects\\data_sepsis\\joint_sepsis.rds") 
```

## Last Observation Carried Forward imputation

```{r}
numeric_vars <- sepsis %>% 
  select(where(is.numeric)) %>% 
  names()

LOCF_sepsis <- sepsis %>%
  group_by(study_patient_id) %>%
  mutate(across(all_of(numeric_vars), ~ zoo::na.locf(.x, na.rm = FALSE))) %>%
  ungroup() 
  
```

## Dataset preparation for modelling

```{r}
prep_sepsis <- LOCF_sepsis %>%
  filter(time_point %in% c(0, 3)) %>% 
  select(study_patient_id, time_point, combined_efferon, age, sex, SOFA, NLR, 
         where(is.numeric), -c(weight, height, BMI, NEU, LYM, APACHE2)) %>% 
  mutate(IL_6 = if_else(IL_6 == 0, 3, IL_6)) %>% 
  mutate(PCT = if_else(PCT == 0, 0.1, PCT))

```

```{r}
prep_sepsis_long <- prep_sepsis %>%
  pivot_longer(
    cols = SOFA:IL_6, 
    names_to = "feature",
    values_to = "value"   
  ) %>%
  select(
    study_patient_id, 
    time_point,
    feature,
    value,
    age,
    sex, 
    combined_efferon,
    SI_CISCIO
  ) %>% 
  mutate(value = round(as.numeric(value), 3))
```

```{r}
add_covariates <- prep_sepsis %>%  
  select(
    study_patient_id,
    time_point,
    SOFA,
    NLR) %>% 
 group_by(study_patient_id) %>% 
  mutate(
    SOFA_0 = first(SOFA[time_point == 0]),
    NLR_0 = first(NLR[time_point == 0])  
  ) %>%
  ungroup() %>% 
  select(-c(SOFA, NLR, time_point))
```

```{r, warning = FALSE}
prep_sepsis_wide <- prep_sepsis_long %>%
  group_by(study_patient_id) %>% 
  pivot_wider(
    names_from = time_point,          
    values_from = value,              
    names_prefix = "value_"           
  )  %>% 
  ungroup() %>% 
  left_join(add_covariates, by = "study_patient_id") %>% 
  select(
    study_patient_id,                
    feature,
    value_0,
    value_3,
    age,
    sex,
    combined_efferon,
    SI_CISCIO,
    SOFA_0,
    NLR_0) %>% 
  distinct()
  
```

```{r}
vis_patients_to_exclude <- prep_sepsis_wide %>%
  filter(feature == "VIS2020", value_0 == 0) %>%
  distinct(study_patient_id)

prep_sepsis_wide <- prep_sepsis_wide %>%
  anti_join(vis_patients_to_exclude, by = "study_patient_id") 
```


## Initial models diagnostics

```{r}
models <- prep_sepsis_wide %>% 
  group_by(feature) %>% 
  mutate(
    value_3_std = (value_3 - mean(value_3, na.rm = TRUE)) / sd(value_3, na.rm = TRUE) # Standardization
  ) %>%
  nest() %>%
  mutate(
    model = map2(
      feature,
      data,
      ~ if (.x %in% c("SOFA", "NLR")) {  
          lm(value_3 ~ value_0 + combined_efferon + age + sex + SI_CISCIO, data = .y) 
        } else {
          lm(value_3 ~ value_0 + combined_efferon + SOFA_0 + NLR_0 + age + sex+ SI_CISCIO, data = .y)  
        }
    ),
    tidy_model = map(model, tidy),
    formula = map_chr(model, ~ paste(deparse(formula(.x)), collapse = " ")),
    diagnostic_plots = map2(model, feature, ~ {
      autoplot(.x, which = 1:4, ncol = 2) +  
        ggtitle(paste("Variable:", .y)) +
        theme_minimal()
    })
  )

models %>%
  select(feature, tidy_model) %>%
  unnest(cols = tidy_model) %>%
  print()

models %>%
  pull(diagnostic_plots) %>%
  walk(print)

```

### Checking models formulas

```{r}
models %>%
  select(feature, formula) %>%
  print()

```

### Log-transform variables 

```{r}

features <- c( "IL_6", "PCT", "aPTT", "bilirubin_total", "creatinine", "lactate")


log_models <- prep_sepsis_wide %>%
  filter(feature %in% features) %>%
  group_by(feature) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(I(log2(.x$value_3)) ~ I(log2(.x$value_0)) + combined_efferon + SOFA_0 + NLR_0 + age + sex + SI_CISCIO, data = .x)),
    tidy_model = map(model, tidy), 
    diagnostics = map(model, autoplot), 
    formula = map_chr(model, ~ paste(deparse(formula(.x)), collapse = " ")) 
  )
log_models %>%
  select(feature, tidy_model) %>%
  unnest(cols = tidy_model) %>%
  print() 

log_models %>%
  mutate(
    plots = map2(model, feature, ~ {
      p <- autoplot(.x, which = 1:4, ncol = 2) +
        ggtitle(paste("Variable:", .y)) +
        theme_minimal()
      print(p)  
      return(p)
    })
  )


log_models %>%
  select(feature, formula) %>%
  print()
```

## Joint models diagnostics

```{r}
  
log_models <- prep_sepsis_wide %>%
  group_by(feature) %>%
  mutate(
    value_3_standardized = ifelse(
      feature %in% features, 
      value_3, 
      (value_3 - mean(value_3, na.rm = TRUE)) / sd(value_3, na.rm = TRUE)
    )
  ) %>%
  nest() %>%
  mutate(
    model = map2(
      feature,
      data,
      ~ if (.x %in% features) {
        
          lm(I(log2(value_3)) ~ I(log2(value_0)) + combined_efferon + SOFA_0 + NLR_0 + age + sex + SI_CISCIO, data = .y)
        } else if (.x %in% c("SOFA", "NLR")) {
          lm(value_3_standardized ~ value_0 + combined_efferon + age + sex + SI_CISCIO, data = .y)
        } else {
          lm(value_3_standardized ~ value_0 + combined_efferon + SOFA_0 + NLR_0 + age + sex + SI_CISCIO, data = .y)
        }
    ),
    tidy_model = map(model, tidy), 
    diagnostics = map(model, autoplot),  
    formula = map_chr(model, ~ paste(deparse(formula(.x)), collapse = " "))  
  )

log_models %>%
  select(feature, tidy_model) %>%
  unnest(cols = tidy_model) %>%
  print()

log_models %>%
  mutate(
    plots = map2(model, feature, ~ {
      p <- autoplot(.x, which = 1:4, ncol = 2) +
        ggtitle(paste("Variable:", .y)) +
        theme_minimal()
      print(p)
      return(p) 
    })
  )
```
### Extracting summary model formulas

```{r}
log_models %>%
  select(feature, formula) %>%
  print()
```

## A summary table for treatment regression coefficients with heteroscedasticity and multiplicity adjustments

```{r, warning = FALSE}
log_models_corrected <- log_models %>%
  mutate(
    corrected_model = map(model, ~ coeftest(.x, vcov. = vcovHC(.x, type = "HC1"))), 
    tidy_corrected = map(corrected_model, ~ broom::tidy(.x, conf.int = TRUE))  
  )

log_models_corrected_table <- log_models_corrected  %>% 
  select(feature, tidy_corrected)  %>% 
    unnest() %>% 
  filter(term == "combined_efferonEfferon") %>% 
  mutate(
    feature = ifelse(feature %in% features, paste0("log2(", feature, ")"), feature)
  ) %>% 
  ungroup() %>% 
  mutate(
    p.adj.BH = p.adjust(p.value, method = "BH")) %>% 
  mutate(across(where(is.numeric), ~ round(., 3))) %>% 
  mutate(significance = case_when(p.value < 0.001 ~ "***", p.value < 0.01 ~ "**", p.value < 0.05 ~ "*", TRUE ~ ""))

datatable(log_models_corrected_table, options = list(pageLength = 10))
```

### Extracting log-transformed variables:

```{r}
log_data <- log_models_corrected_table %>%
  filter(feature %in% paste0("log2(", features, ")"))
```

### Extracting non-log-transformed variables:

```{r}
non_log_data <- log_models_corrected_table %>%
  filter(!feature %in% paste0("log2(", features, ")"))
```


## Forest plot for variables with log-transformation

```{r}
label_text_log <- cbind(
  Variable = log_data$feature,
  Estimate = sprintf("%.2f", log_data$estimate),
  CI = sprintf("[%.2f, %.2f]", log_data$conf.low, log_data$conf.high),
  Significance = log_data$significance)

forestplot(
  labeltext = label_text_log,
  mean = log_data$estimate,
  lower = log_data$conf.low,
  upper = log_data$conf.high,
  is.summary = FALSE,
  xlab = "Exponentiated Coefficient (95% CI)",
  zero = 0, boxsize = 0.3,
  col = fpColors(box = "blue", line = "darkblue", zero = "red"),
  title = "Forest plot (variables with log-transformation)",
  txt_gp = fpTxtGp(xlab = gpar(fontsize = 20), ticks = gpar(fontsize = 20)),
  clip = c(-2, 2)  
)
```


## Forest plot for variables without log-transformation

```{r}
label_text_non_log <- cbind(
  Variable = non_log_data$feature,
  Estimate = sprintf("%.2f", non_log_data$estimate),
  CI = sprintf("[%.2f, %.2f]", non_log_data$conf.low, non_log_data$conf.high),
  Significance = non_log_data$significance)

forestplot(
  labeltext = label_text_non_log,
  mean = non_log_data$estimate,
  lower = non_log_data$conf.low,
  upper = non_log_data$conf.high,
  is.summary = FALSE,
  xlab = "Standardized Coefficient Estimate (95% CI)",
  zero = 0, boxsize = 0.3,
  col = fpColors(box = "blue", line = "darkblue", zero = "red"),
  title = "Forest plot (variables without log-transformation)",
  txt_gp = fpTxtGp(xlab = gpar(fontsize = 20), ticks = gpar(fontsize = 20)),
  clip = c(-2, 2)  
)
```




